Relict deciduous oaks show high resilience to drought despite the effect of land-use legacies

### 2019-05-21 man/ms.Rmd
update

---
title: "Relict deciduous oaks show high resilience to drought despite the effect of land-use legacies"
author: "AJ Perez-Luque; G. Gea-Izquierdo; Zamora, R. ..."
output:
  word_document:
    fig_caption: yes
    fig_height: 6
    fig_width: 6
    highlight: null
    reference_docx: ../templates/template_ms.docx
bibliography: ../refs/references.bib
csl: ../refs/ecology.csl
---

@@

* Despite being considered relict stands at the dry rear-edge for the species, they showed high resilience in response to drought.

Yet, the high sensitivity to water availability expressed and positive growth trends only in the coldest and more humid stands suggest divergent response to a changing climate along very short climatic gradients and suggest that dynamics (i.e. not only adult performance but also regeneration) of those stands located at the lowest elevations, hence the altitudinal rear-edge, need to be monitored under future expected drier and warmer climatic scenarios and, particularly, if high management pressure remains.

@@

The analysis of tree-growth dynamics revealed suppression and release events that are consistent with socieconomic dynamic, derived from the review of historical documents. Recently researchs have combined information from both tree-ring and remote-sensing to analyze the resilience of forest to drought [*e.g.* @Gazol2018]. However our approach, which includes also an exhaustive review of historical documents, helps to disentangle the impacts of climatic and land use for forest systems. In this sense our results highligth the importance of landuse legacies for highly transformed Mediterranean systemas [@Navarro2013].

Growth projections have forecasted a decrease in productivity for *Q. pyrenaica* that would increase vulnerability of this species to climate warming at the dry edges locations [@GeaIzquierdo2013]. However, the positive trens for both primary (*i.e.* greenness) and secondary growth (BAI) observed ar our study site and the high resilience values to drought showed at long term scale, seem to suggest that Sierra Nevada constitutes still a refuge for some species. This is relevant for tree species with higher sensitivity to climate change, such us *Q. pyrenaica* (García-Valdes et al. 2013; Benito-Garzón et al. 2008), no solo por la conservación de la especie en sí, sino por los servicios ecosistémicos que estos bosques tan sigulares ofrecen.

En este sentido, los futuros trabajos deberan ir en ...

VAS POR AQUI

En estas zonas refugio en particular parece tener mucha importancia el abandono de las actividades tradicionales, que por un lado y de acuerdo a nuestros resultados parecen estar condicionando el crecimiento

Future work:

C5 Servicios ecosistémicos

Aumento de crecimiento 
Refugio para biodivesidad 
Pero ojo ---> estancamiento y/o mortalidad?? Necesidad de prácticas de gestión que aseguren un crecimiento viable, promoviendo la regeneración sexual (evitar el estancamiento) para asi mantener los 
La tendencia creciente del crecimiento en el crecimiento observada en los últimos años podría relacionarse con un abandono de las practicas tradicionales en esta region de montaña, así como 
el abandono de las prácticas tradicionales así como por el caracter de re

Además las diferencias en el crecimiento entre los dos sitios mas cercanos así como lo de la resiliencia, van en la línea de lo expresado por Vilá referente a lo de microrefugios, por lo que la gestión de estas masas, situadas en su rear edge, además ha de analizarse teniendo en cuenta este aspecto.

El abandono de las masas de Q. pyrenaica ha provocado un estado general de degradación, que puede tener consecu
La reducción de la presión antrópica durante las últimas décadas ha provocado que estas bosques presenten un avanzado estado de degradación

After at least five decades of abandonment, old Cs of Q. pyrenaica in Spain present a general state of degradation

In this work, molecular tools have been successfully applied to assess levels of genetic diversity in Cs, rejecting one of the most widespread and preconceived idea in Spanish silviculture: the decline in genetic diversity promoted by historical coppicing in Q. pyrenaica.

En este sentido, sabemos que estos robledales han estado sometido a muchos ciclos de coppiccing, lo que podrían haber reducido su diversidad genética, y por tanto su resiliencia. Pero varios trabajos [@Valbuena2013; @Valbuena2017] han encontrado que esta especie en su límite de distribución sur, muestra unos altos niveles de resiliencia (en este caso genética).

## References

>>>>

---
title: "Relict deciduous oaks show high resilience to drought despite the effect of land-use legacies"
author: "A.J. Perez-Luque; G. Gea-Izquierdo and Zamora, R."
output:
  word_document:
    fig_caption: yes
    fig_height: 6
    fig_width: 6
    highlight: null
    reference_docx: ../templates/template_ms.docx
bibliography: ../refs/references.bib
csl: ../refs/ecology.csl
---

@@

* Despite being considered relict stands at the dry rear-edge for the species, they showed high resilience in response to drought. Yet, the high sensitivity to water availability expressed and positive growth trends only in the coldest and more humid stands suggest divergent response to a changing climate along very short climatic gradients and suggest that dynamics (i.e. not only adult performance but also regeneration) of those stands located at the lowest elevations, hence the altitudinal rear-edge, need to be monitored under future expected drier and warmer climatic scenarios and, particularly, if high management pressure remains.

@@

The analysis of tree-growth dynamics revealed suppression and release events that are consistent with socieconomic dynamic, derived from the review of historical documents. Recently works have combined information from both tree-ring and remote-sensing to analyze the resilience of forest to drought [*e.g.* @Gazol2018]. The approach presented here, which also includes an exhaustive review of historical documents, help to identify the response of vegetation to climate and to land use changes. In this sense our results highligth the importance of landuse legacies for highly transformed Mediterranean systemas [@Navarro2013].

Growth projections have forecasted a decrease in productivity for *Q. pyrenaica* that would increase vulnerability of this species to climate warming at the dry edges locations [@GeaIzquierdo2013]. However, the positive trends for both primary (*i.e.* greenness) and secondary growth (BAI) observed ar our study site and the high resilience values to drought showed at long term scale, seem to suggest that Sierra Nevada constitutes still a refuge for some species. This is relevant for tree species with higher sensitivity to climate change, such us *Q. pyrenaica* (García-Valdes et al. 2013; Benito-Garzón et al. 2008), not only for conservation per se of this species, but for all ecosystem services that these singular forests offer at Mediterranean mountain.

The abandonment of traditional activities along with the increase in temperature since the 70s, could be explaining the positive growth trend observed. This trend and the increase of tree-cover observed in the last decades in this mountain region [@CamachoOlmedo2002], would enhance the carbon sequestration of this forest promoting its role as carbon sink. However, the abandonment of traditional management has provoked a general state of degradation of this forest formation [@Canellas2004; @Corcuera2006]. In this sense, es necesario analizar con mas detalle como se verán afectados los servicios ecosistémicos que ofrecen estos robledales en su rear edge en un contexto de cambio global, identificando claramente los impactos del clima y del land-use change en la provisión de servicios ecosistémicos.

# Acknowledgments

We would like to thank Sammy Lucy Behle, M. Suarez-Muñoz and F.J. Bonet-García for their assistence in the field monitoring. A. Reyes and J. Blanco helped in the filtering of satellite data. We also are very thankful to F.J. Bonet-García for its valuable comments in an earlier version of the manuscript. Environment Department of Andalusian Regional Government and Director of Sierra Nevada National and Natural Park provided permission for field work. This research work was conducted in the collaborative framework of the “Sierra Nevada Global Change Observatory” Project from the Environment Department of Andalusian Regional Government (with the support of European Union FEDER Project). We also thank to LIFE-ADAPTAMED (LIFE14 CCA/ES/000612): *Protection of key ecosystem services by adaptive management of Climate Change endangered Mediterranean socioecosystems* and H2020 project *ECOPOTENTIAL: Improving future ecosystem benefits through earth observations* (http://www.ecopotential-project.eu/) (No 641762), for their funding support.

# References
